Fractional Types

### 2013-02-03 frac.tex
sec 2

\todo{brief explanation}

\begin{definition}[Primitive {{langRev}} Combinators]
\label{def:primitives-langRev}

%subcode{bnf} include main
%! columnStyle = rrcll
%  zeroe :&  0 + b &<->& b &: zeroi
%  swap+ :&  b1 + b2 &<->& b2 + b1 &: swap+
%  assocl+ :&  b1 + (b2 + b3) &<->& (b1 + b2) + b3 &: assocr+
%  unite :&  1 * b &<->& b &: uniti
%  swap* :&  b1 * b2 &<->& b2 * b1 &: swap*
%  assocl* :&  b1 * (b2 * b3) &<->& (b1 * b2) * b3 &: assocr*
%  dist0 :&  0 *b &<->& 0  &: factor0
%  dist :& (b1 + b2) * b3 &<->& (b1 * b3) + (b2 * b3) &: factor

\end{definition}

Each line of this table is to be read as the definition of one or two
combinators. For example, corresponding to the \textit{identity of
  multiplication} ({{1*b<->b}}) we have the two combinators
{{unite:1*b<->b}} (reading the isomorphism from left to right) and
{{uniti:b<->1*b}} (reading the isomorphism from right to left). These
combinators are inverses of each other.  Each of the two cases of
commutativity defines one combinator that is its own inverse.

Now that we have primitive combinators we need some means of composing
them. We construct the composition combinators out of the congruence
closure of basic isomorphisms. Thus we have program constructs that
witness reflexivity {{id}}, symmetry {{sym}}, transitivity {{(;)}},
and two parallel composition combinators, one for sums {{(+)}} and one
for pairs {{(*)}}.

\begin{definition}[Composition in {{langRev}}]
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%----
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
\end{definition}

The semantics of {{langRev}} is given using two mutually recursive
interpreters: one going forward and one going backwards. The use of {{sym}}
switches control from one evaluator to the other. We will not present the
semantics because we generalize in the next section.

\begin{proposition}[Termination] Evaluation of well-typed combinators
  always terminates. In other words, for {{c : b1 <-> b2}} and 
  {{v1 : b1}} there exists {{v2 : b2}} such that {{evalF c v1 = v2}}. 
  Conversely, if {{v2 : b2}} there exists {{v1 : b1}} such that 
  {{evalB c v2 = v1}}.
\end{proposition}

\begin{proposition}[Logical Reversibility]
\label{prop:logrev}
For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
{{v2 : b2}} we have: {{evalF c v1 = v2}} iff {{evalB c v2 = v1}}.
\end{proposition}

\todo{mention that we have extended this with trace and recursive types but
  that we are going to explore something orthogonal}

@@

\begin{definition}[Syntax of {{langRevT}}]
\label{def:langRevT}
The syntax of types, values, and combinators, is:
%subcode{bnf} include main
% natural numbers, n 
% value types, b ::= 0 | 1 | b+b | b*b | 1/b | bool | nat 
% values, v ::= () | left v | right v | (v,v) | 1/v | true | false | n
%
% combinator types, t ::= b <-> b
% isomorphisms, iso ::= zeroe | zeroi | swap+ | assocl+ | assocr+ 
%                   &|& unite | uniti | swap* | assocl* | assocr* 
%                   &|& dist0 | factor0 | dist | factor 
%                   &|& foldB | unfoldB | foldN | unfoldN
%                   &|& eta*_b | eps*_b
% combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
%                   &|& trace+ c | trace* c
\end{definition}

@@

\begin{definition}[Denotation of Types {{ [[b]] }}]
\label{chx:def:denot}
Each type denotes a set of values as follows:
\[\begin{array}{rcl}
\llbracket 0 \rrbracket &=& \emptyset \\
\llbracket 1 \rrbracket &=& \{ () \} \\
\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \}
           \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, 
           v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} \\
\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
\llbracket \mathit{nat} \rrbracket &=& \{ 0, 1, 2, \ldots \} 
\end{array}\]
\end{definition}

@@

%subcode{opsem} include main
%! columnStyle = rcl
% evalF zeroe (right v) &=& {[ v ]}
% evalF zeroi v &=& {[ right v ]}
% evalF swap+ (left v) &=& {[ right v ]}
% evalF swap+ (right v) &=& {[ left v ]}
% evalF assocl+ (left v) &=& {[ left (left v) ]}
% evalF assocl+ (right (left v)) &=& {[ left (right v) ]}
% evalF assocl+ (right (right v)) &=& {[ right v ]}
% evalF assocr+ (left (left v)) &=& {[ left v ]}
% evalF assocr+ (left (right v)) &=& {[ right (left v) ]}
% evalF assocr+ (right v) &=& {[ right (right v) ]}
% evalF unite ((), v) &=& {[ v ]}
% evalF uniti v &=& {[ ((), v) ]}
% evalF swap* (v1, v2) &=& {[ (v2, v1) ]}
% evalF assocl* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
% evalF assocr* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
% evalF dist (left v1, v3) &=& {[ left (v1, v3) ]}
% evalF dist (right v2, v3) &=& {[ right (v2, v3) ]}
% evalF factor (left (v1, v3)) &=& {[ (left v1, v3) ]}
% evalF factor (right (v2, v3)) &=& {[ (right v2, v3) ]}
% evalF id v &=& {[ v ]}
% evalF (sym c) v &=& evalB c v
% evalF (c1(;)c2) v &=& Union ~{[ evalF c2 v' ~|~ v' in (evalF c1 v) ]}
% evalF (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalF c1 v1) ]}
% evalF (c1 (+) c2) (right v2) &=& {[ left v' ~|~ v' in (evalF c2 v2) ]}
% evalF (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalF c1 v1), v4 in (evalF c2 v2) ]}
% evalF (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalF c (v,v2)) ]}
% evalF ~eta*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
% evalF ~eps*_b (1/v,v) &=& {[ () ]} 
% evalF ~eps*_b (1/v1,v2) &=& emptyset, where v1 /= v2

\paragraph*{Backwards Evaluator:}

%subcode{opsem} include main
%! columnStyle = rcl
% evalB zeroe v &=& {[ right v ]}
% evalB zeroi (right v) &=& {[ v ]}
% evalB swap+ (left v) &=& {[ right v ]}
% evalB swap+ (right v) &=& {[ left v ]}
% evalB assocl+ (left (left v)) &=& {[ left v ]}
% evalB assocl+ (left (right v)) &=& {[ right (left v) ]}
% evalB assocl+ (right v) &=& {[ right (right v) ]}
% evalB assocr+ (left v) &=& {[ left (left v) ]}
% evalB assocr+ (right (left v)) &=& {[ left (right v) ]}
% evalB assocr+ (right (right v)) &=& {[ right v ]}
% evalB unite v &=& {[ ((), v) ]}
% evalB uniti ((), v) &=& {[ v ]}
% evalB swap* (v1, v2) &=& {[ (v2, v1) ]}
% evalB assocl* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
% evalB assocr* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
% evalB dist (left (v1, v3)) &=& {[ (left v1, v3) ]}
% evalB dist (right (v2, v3)) &=& {[ (right v2, v3) ]}
% evalB factor (left v1, v3) &=& {[ left (v1, v3) ]}
% evalB factor (right v2, v3) &=& {[ right (v2, v3) ]}
% evalB id v &=& {[ v ]}
% evalB (sym c) v &=& evalF c v
% evalB (c1(;)c2) v &=& Union ~{[ evalB c1 v' ~|~ v' in (evalB c2 v) ]}
% evalB (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalB c1 v1) ]}
% evalB (c1 (+) c2) (right v2) &=& {[ right v' ~|~ v' in (evalB c2 v2) ]}
% evalB (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalB c1 v1), v4 in (evalB c2 v2) ]}
% evalB (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalB c (v,v2)) ]}
% evalB ~eta*_b (1/v,v) &=& {[ () ]} 
% evalB ~eta*_b (1/v1,v2) &=& emptyset, where v1 /= v2
% evalB ~eps*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
\end{definition}

>>>>

We review our language {{langRev}} providing the necessary background and
context for our higher-order extension.\footnote{The presentation in this
  section focuses on the simplest version of {{langRev}}. Other versions
  include the empty type, recursive types, and trace opeators but these
  extensions are orthogonal to the higher-order extension emphasized in this
  paper.} The terms of {{Pi}} are not classical values and functions; rather,
the terms are isomorphism witnesses.  In other words, the terms of {{Pi}} are
proofs that certain ``shapes of values'' are isomorphic.  And, in classical
Curry-Howard fashion, our operational semantics shows how these proofs can be
directly interpreted as actions on ordinary values which effect this shape
transformation. Of course, ``shapes of values'' are very familiar already:
they are usually called \emph{types}.  But usually one designs a type system
as a method of classifying terms, with the eventual purpose to show that
certain properties of well-typed terms hold, such as safety.  Our approach is
different: we start from a type system, and then present a term language
which naturally inhabits these types, along with an appropriate operational
semantics.

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
and these are classified by ordinary types
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b

Types include the unit type {{1}}, sum types {{b1+b2}}, and products types
{{b1*b2}}.  Values includes {{()}} which is the only value of type {{1}},
{{left v}} and {{right v}} which inject {{v}} into a sum type, and
{{(v1,v2)}} which builds a value of product type. But these values should be
regarded as largely ancilliary.  In particular we do not treat the above
values as first-class citizens.  They only occur when we want to observe the
effect of an isomorphism.

\paragraph*{Isomorphisms.} The terms of {{Pi}} are witnesses to type
isomorphisms.  They have (iso) types {{b <-> b}}.  
Specifically, they are witnesses to the following type isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
Each line of the above table introduces a pair of dual constants%
\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness the type
isomorphism in the middle.  These are the base (non-reducible) terms of the
second, principal level of {{Pi}}.

Note how the above has two readings: first as a set of typing relations for a
set of constants. If these axioms are seen as universally quantified,
orientable statements, they also induce transformations of the (traditional)
values.  The (categorical) intuition here is that these axioms have
computational content because they witness isomorphisms rather than merely
stating an extensional equality.

For example {{ swap* }} induces a function (akin to):
%subcode{code}
%! language = Haskell
% swapProd (a,b) = (b,a)

The isomorphisms are extended to form a congruence relation by adding the
following constructors that witness equivalence and compatible closure:
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4

The syntax is overloaded: we use the same symbol at the value-type level
and at the isomorphism-type level for denoting sums and products.  Hopefully
this will not cause undue confusion.

It is important to note that ``values'' and ``isomorphisms'' are 
completely separate syntactic categories which do not intermix.  The 
semantics of the language come when these are made to interact at
the ``top level'', a new syntactic category:
%subcode{bnf} include main
% top level term, l ::= c v
We refer to this as \emph{application}.

\noindent
To summarize, the syntax of {{Pi}} is given as follows.

\begin{definition}{(Syntax of {{Pi}})}
\label{def:Pi}
We collect our types, values, and isomorphisms, to get the full language
definition.
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%     &|& identl* | identr* 
%     &|& swap* | assocl* | assocr* 
%     &|& dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top level term, l ::= c v
\end{definition}

The semantics of {{langRev}} is given using two mutually recursive
interpreters: one going forward and one going backwards. The use of {{sym}}
switches control from one evaluator to the other. We will not present the
semantics because we present a more general version in the next section. We
state without proof that the evaluation of well-typed combinators always
terminates and that {{langRev}} is logically reversible, i.e., that for all
combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have
the forward evaluation of {{c v1}} produces {{v2}} iff the backwards
evaluation of {{c v2}} produces {{v1}}.

@@

\begin{definition}[Syntax of {{langRevT}}]
\label{def:langRevT}
The syntax of types, values, and combinators, is:
%subcode{bnf} include main
% natural numbers, n 
% value types, b ::= 1 | b+b | b*b | 1/b | bool 
% values, v ::= () | left v | right v | (v,v) | 1/v | true | false | n
%
% combinator types, t ::= b <-> b
% isomorphisms, iso ::= swap+ | assocl+ | assocr+ 
%                   &|& unite | uniti | swap* | assocl* | assocr* 
%                   &|& dist | factor 
%                   &|& foldB | unfoldB 
%                   &|& eta*_b | eps*_b
% combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
\end{definition}

@@

\begin{definition}[Denotation of Types {{ [[b]] }}]
\label{chx:def:denot}
Each type denotes a set of values as follows:
\[\begin{array}{rcl}
\llbracket 1 \rrbracket &=& \{ () \} \\
\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \}
           \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, 
           v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} 
\end{array}\]
\end{definition}

@@

%subcode{opsem} include main
%! columnStyle = rcl
% evalF swap+ (left v) &=& {[ right v ]}
% evalF swap+ (right v) &=& {[ left v ]}
% evalF assocl+ (left v) &=& {[ left (left v) ]}
% evalF assocl+ (right (left v)) &=& {[ left (right v) ]}
% evalF assocl+ (right (right v)) &=& {[ right v ]}
% evalF assocr+ (left (left v)) &=& {[ left v ]}
% evalF assocr+ (left (right v)) &=& {[ right (left v) ]}
% evalF assocr+ (right v) &=& {[ right (right v) ]}
% evalF unite ((), v) &=& {[ v ]}
% evalF uniti v &=& {[ ((), v) ]}
% evalF swap* (v1, v2) &=& {[ (v2, v1) ]}
% evalF assocl* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
% evalF assocr* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
% evalF dist (left v1, v3) &=& {[ left (v1, v3) ]}
% evalF dist (right v2, v3) &=& {[ right (v2, v3) ]}
% evalF factor (left (v1, v3)) &=& {[ (left v1, v3) ]}
% evalF factor (right (v2, v3)) &=& {[ (right v2, v3) ]}
% evalF id v &=& {[ v ]}
% evalF (sym c) v &=& evalB c v
% evalF (c1(;)c2) v &=& Union ~{[ evalF c2 v' ~|~ v' in (evalF c1 v) ]}
% evalF (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalF c1 v1) ]}
% evalF (c1 (+) c2) (right v2) &=& {[ left v' ~|~ v' in (evalF c2 v2) ]}
% evalF (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalF c1 v1), v4 in (evalF c2 v2) ]}
% evalF ~eta*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
% evalF ~eps*_b (1/v,v) &=& {[ () ]} 
% evalF ~eps*_b (1/v1,v2) &=& emptyset, where v1 /= v2
\paragraph*{Backwards Evaluator:}

%subcode{opsem} include main
%! columnStyle = rcl
% evalB swap+ (left v) &=& {[ right v ]}
% evalB swap+ (right v) &=& {[ left v ]}
% evalB assocl+ (left (left v)) &=& {[ left v ]}
% evalB assocl+ (left (right v)) &=& {[ right (left v) ]}
% evalB assocl+ (right v) &=& {[ right (right v) ]}
% evalB assocr+ (left v) &=& {[ left (left v) ]}
% evalB assocr+ (right (left v)) &=& {[ left (right v) ]}
% evalB assocr+ (right (right v)) &=& {[ right v ]}
% evalB unite v &=& {[ ((), v) ]}
% evalB uniti ((), v) &=& {[ v ]}
% evalB swap* (v1, v2) &=& {[ (v2, v1) ]}
% evalB assocl* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
% evalB assocr* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
% evalB dist (left (v1, v3)) &=& {[ (left v1, v3) ]}
% evalB dist (right (v2, v3)) &=& {[ (right v2, v3) ]}
% evalB factor (left v1, v3) &=& {[ left (v1, v3) ]}
% evalB factor (right v2, v3) &=& {[ right (v2, v3) ]}
% evalB id v &=& {[ v ]}
% evalB (sym c) v &=& evalF c v
% evalB (c1(;)c2) v &=& Union ~{[ evalB c1 v' ~|~ v' in (evalB c2 v) ]}
% evalB (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalB c1 v1) ]}
% evalB (c1 (+) c2) (right v2) &=& {[ right v' ~|~ v' in (evalB c2 v2) ]}
% evalB (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalB c1 v1), v4 in (evalB c2 v2) ]}
% evalB ~eta*_b (1/v,v) &=& {[ () ]} 
% evalB ~eta*_b (1/v1,v2) &=& emptyset, where v1 /= v2
% evalB ~eps*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
\end{definition}
